---
title: The hypothesis of a ‘core’ community receives poor support when confronted
  with simulated and empirical data
output:
  html_document:
    df_print: paged
---

## Maya Gans and Gordon Custer
## University of Wyoming
## December 3, 2018


# Load Packages

```{r, results='hide'}
require(tidyverse)
require(matrixStats)
require(gridExtra)
require(parallelDist)
```


# Import Data

Import OTU tables: OTUs as rows, sample replicates as columns. Here we upload two different datasets

```{r, warning = FALSE}
human <- readRDS("../data/human.RDS")
arabadopsis <- readRDS("../data/arabadopsis.RDS")
```


# Mean, Variance, Covariance

In this section we calculate the mean, variance and CV of each OTU.  This will be used downstream to see if there is correlation between CV/Mean/Variance and the OTUs defined as core.  The end product of this section is a dataframe with the Mean, Variance and CV for each OTU. 

```{r}
summarize_OTUS <- function(otu_table) {
  otu_table %>%
    pivot_longer(-1) %>% 
    group_by(X) %>% 
    summarise(Mean = mean(value), Variance = var(value), CV = Variance/Mean)
}
```


# Four Methods for Looking at the Core-Non Core Divide 

### Proportion of reads

This method assigns taxa to the core if they are in the top X% of reads. Taxa are ranked in abudnace and the cumulative sum is recoreded. Any taxa which appears before some cutoff percentage is included in the core. In this example, a taxa will be assigned to the core if they account for the first 75% of the reads

```{r}
prop_reads <- function(otu_table) {
  otu_table %>%
  pivot_longer(-X) %>%
  group_by(X) %>%
  summarise(otuappearance = sum(value)) %>%
  arrange(-otuappearance) %>%
  mutate(s = sum(otuappearance),
         prop = otuappearance/s,
         cumsum = cumsum(prop)) %>%
  filter(cumsum <= 0.75) %>%
  pull(X)
}
```

### Proportion of replicates method

This method assigns taxa to the core based upon the number of sites it is present in. In this example we assign core membership when the taxa's abundance is atleast 10 fold the number of sites. This method accounts for abundance as a function of sites. 

```{r}
prop_reps <- function(otu_table) {
  otu_table %>%
    pivot_longer(-1) %>%
    group_by(X) %>%
    summarise(abundance = sum(value)) %>%
    filter(abundance > 10*ncol(otu_table)) %>%
    pull(X)
}
```

### Proportion of reads and replicates

This method assigns taxa to the core if they account for some proportion of the total reads for the sequencing run and if they are preseant in atleast x% of the total number of replicates. In this example, a core taxa must account for 0.01% of the total reads for the entire otu table and be present in at least 50% of sites.

```{r}
prop_reads_and_reps <- function(otu_table) {
  otu_table %>%
    pivot_longer(-X) %>%
    group_by(X) %>%
    summarise(rowsum = sum(value)) %>%
    mutate(s = sum(rowsum)) %>%
    left_join((otu_table %>% pivot_longer(-X)), by = "X") %>%
    filter(value > 0) %>%
    group_by(X) %>%
    summarise(
      num_sites = n(),
      N = ncol(otu_table) - 1) %>%
    filter(num_sites >= N * 0.5) %>%
    pull(X)
}
```

### Hard cut offs

This method assigns taxa to the core if they are present in more than a pre-determined number of sites and have a total abudance greater than a pre-determined number of reads. In our example we set the minimum number of sites to 5 and the minimum number of reads to 25. Here we use the hard cut off described in Lundberg (2012), but realize this is any threshhold. 

```{r}
hard_cutoff <- function(otu_table) {
  otu_table %>%
    pivot_longer(-X) %>%
    select(-name) %>%
    filter(value > 25) %>%
    group_by(X) %>%
    summarise(count = n()) %>%
    filter(count >= 5) %>%
    pull(X)
}
```


# Combine Methods

Make a dataframe with all observed taxa, their inclusion to the core by method (deliniated as a 1 or 0), the mean, variance, and coefficient of variation.

```{r}
combine_methods <- function(otu_table) {
summarize_OTUS(otu_table) %>%
  mutate(
    `Proportion of Sequence Reads` = X %in% prop_reads(otu_table),
    `Proprortion of Sequence Reads and Replicates` = X %in% prop_reads_and_reps(otu_table),
    `Hard Cut Off` = X %in% hard_cutoff(otu_table),
    `Proportion of Sequence Replicates` = X %in% prop_reps(otu_table)) %>%
    mutate_if(is.logical, as.numeric) %>%
    pivot_longer(cols = 5:8) %>%
    mutate(name = factor(name, levels = c("Proportion of Sequence Reads",
                                          "Proportion of Sequence Replicates",
                                          "Proprortion of Sequence Reads and Replicates",
                                          "Hard Cut Off")))
}
```

```{r}
combined_arabadopsis <- combine_methods(arabadopsis)
combined_human<- combine_methods(human)
```


# Make a copy of each dataframe as its produced

```{r}
write.csv(combined_arabadopsis, "../data_output/Core_Inclusion_and_Stats_Arabadopsis")
write.csv(combined_human, "../data_output/Core_Inclusion_and_Stats_Human")
```

# Plot

A single plot looking at the 4 different methods of core inclusion.

```{r}
combined_plots <- function(combined_otu_data, low = "#EAE6f3", high = "#432976", legend_title) {
  combined_otu_data %>%
    ggplot() +
    aes(x = log(Mean), y = CV, color = as.factor(value)) +
    geom_hex(bins = 30) +
    scale_color_manual(values = c("lightgray", "black")) +
    scale_fill_gradient(low = low, high = high) +
    theme_bw() +
  facet_grid(.~ name) +
  guides(color = FALSE) + 
  ylab("Coefficent of Variance") +
  labs(fill = legend_title)
}  
```

```{r}
combined_plots(combined_arabadopsis, high = "#2D5062", low = "#E6ECF1", legend_title = "Arabidopsis Taxa Count")
```


# Figure for Manuscript 

Combine the dataframes to facet together

```{r}
all_data_combined <- rbind(
  combined_arabadopsis %>% mutate(Legend = "Arabidopsis Taxa Count"),
  combined_human %>% mutate(Legend = "HMP Taxa Count")
)

figure_2 <- all_data_combined %>%
    ggplot() +
    aes(x = log(Mean), y = CV, color = as.factor(value)) +
    geom_hex(bins = 30) +
    theme_bw() +
    scale_color_manual(values = c("lightgray", "black")) +
    scale_fill_gradient(low = "#EAE6f3", high = "#432976") +
    theme_bw() +
  facet_grid(Legend ~ name) +
  guides(color = FALSE) + 
  ylab("Coefficent of Variance") +
  labs(fill = "Taxa Count")

ggsave(figure_2, "figure_2.png")
```
