---
title: "Simulation of Core"
author: Alex Buerkle, Maya Gans, and Gordon Custer 
output:
  html_notebook:
    code_folding: hide
---

## University of Wyoming
## 4 June 2019

We load `MCMCpack` to draw random samples fron a Dirichlet.
```{r, warning = FALSE}
library(MCMCpack)
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
library(gridExtra)
```

```{r}
# set the structure of our simlulated taxa table (not proritions of taxa, just rows by columns and expected core)
ntaxa<-1000
ncore<-25
nsites<-50
nreads<-2e06
nsims<-250
```

Sets ratio of relative abundance $\pi$ of core to non-core taxa.

$$\frac{\pi \text{ core}}{\pi \text{ non-core}} $$
The relative probability of drawing a member of a core taxon is dependent on how large $\frac{\pi \text{ core}}{\pi \text{ non-core}}$ is. Set intensity prameter $\theta$; this controls how much variance is allowed. 
```{r}
corenoncoreratio<-c(1,2,5,10,25)
p.core<-corenoncoreratio/(ntaxa - ncore + corenoncoreratio*ncore)
p.noncore<-( 1-(p.core*ncore) )/ (ntaxa - ncore)

theta<-c(1, 2, 10, 25, 50) #intensity parameter
```

```{r}
## do a set of simulations across input parameter values and number of replicates ------
## set up the storage of input and output for the 25 simulation conditions
## quantify TP (true positives) and FP (false positives)
sim.df<-data.frame(expand.grid(corenoncoreratio=corenoncoreratio, theta=theta), 
           p.core=rep(p.core, length(theta)), 
           p.noncore=rep(p.noncore, length(theta)),

           meanerrorHardCutTP= numeric(length(p.core) * length(theta)),
           meanerrorPropReadTP = numeric(length(p.core) * length(theta)),
           meanerrorPropRepsTP = numeric(length(p.core) * length(theta)),
           meanerrorPropReadRepsTP = numeric(length(p.core) * length(theta)),
          meanerrorHardCutFP= numeric(length(p.core) * length(theta)),
           meanerrorPropReadFP = numeric(length(p.core) * length(theta)),
           meanerrorPropRepsFP = numeric(length(p.core) * length(theta)),
           meanerrorPropReadRepsFP = numeric(length(p.core) * length(theta))
           )
           #error2.5 = numeric(length(p.core) * length(theta)),
           #error97.5 = numeric(length(p.core) * length(theta)))

#sim.df
```

```{r}
for(trt in 1:(nrow(sim.df))){
    
    # create empty vectors to store output for each method for every simulation
    tmp.err.HardCutTP <-  numeric(nsims)
    tmp.err.PropReadTP <- numeric(nsims)
    tmp.err.PropRepsTP <- numeric(nsims)
    tmp.err.PropReadRepsTP <- numeric(nsims)
    tmp.err.HardCutFP <-  numeric(nsims)
    tmp.err.PropReadFP <- numeric(nsims)
    tmp.err.PropRepsFP <- numeric(nsims)
    tmp.err.PropReadRepsFP <- numeric(nsims)
  for(i in 1:nsims){
    # create a single OTU table for all four methods to be compared against
    replicateSim <- round(rdirichlet(nsites, c(rep(sim.df$p.core[trt] * sim.df$theta[trt], ncore), 
                                               rep(sim.df$p.noncore[trt] * sim.df$theta[trt], ntaxa-ncore))) * nreads)
    
    # find true and false positives
    # true positives are those that meet in the criterion in the first 1:ncore taxa.  The remaining taxa are not part of the core, so any taxa meeting the criterion in that set are false positives
    
    # hard cut off method
    # tmp.err.HardCut[i] <-  sum(colSums((replicateSim >= 25)) >= 5) - ncore
    tmp.err.HardCutTP[i] <-  sum((colSums((replicateSim >= 25)) >= 5)[1:ncore])
    tmp.err.HardCutFP[i] <-  sum((colSums((replicateSim >= 25)) >= 5)[(ncore+1):ntaxa])

    # Prop reads
    #tmp.err.PropRead[i] <- sum(cumsum(sort(colSums(replicateSim)/sum(rowSums(replicateSim))))>=0.25)  - ncore
    tmp.err.PropReadTP[i] <- sum((cumsum(sort(colSums(replicateSim)/sum(rowSums(replicateSim))))>=0.25)[1:ncore])
    tmp.err.PropReadFP[i] <- sum((cumsum(sort(colSums(replicateSim)/sum(rowSums(replicateSim))))>=0.25)[(ncore+1):ntaxa])  
    # Prop Reps
    #tmp.err.PropReps[i] <- sum(colSums(replicateSim)>10*nrow(replicateSim))  - ncore
    tmp.err.PropRepsTP[i] <- sum((colSums(replicateSim)>10*nrow(replicateSim))[1:ncore])  
    tmp.err.PropRepsFP[i] <- sum((colSums(replicateSim)>10*nrow(replicateSim))[(ncore+1):ntaxa])   
    # Thanks Alex
    #tmp.err.PropReadReps[i] <- sum((colSums(replicateSim>0) >= (0.5 * nsites)) & (colSums(replicateSim) >= sum(replicateSim) * 1/1000) )  - ncore
    tmp.err.PropReadRepsTP[i] <- sum(((colSums(replicateSim>0) >= (0.5 * nsites)) & (colSums(replicateSim) >= sum(replicateSim) * 1/1000) )[1:ncore]) 
    tmp.err.PropReadRepsFP[i] <- sum(((colSums(replicateSim>0) >= (0.5 * nsites)) & (colSums(replicateSim) >= sum(replicateSim) * 1/1000) )[(ncore+1):ntaxa]) 
  }
  
  # fill the dataframe from the chunk above with these values for later plotting
  sim.df$meanerrorHardCutTP[trt] <- mean(tmp.err.HardCutTP) / ncore
  sim.df$meanerrorPropReadTP[trt] <- mean(tmp.err.PropReadTP) / ncore
  sim.df$meanerrorPropRepsTP[trt] <- mean(tmp.err.PropRepsTP) / ncore
  sim.df$meanerrorPropReadRepsTP[trt] <- mean(tmp.err.PropReadRepsTP) / ncore
  
  sim.df$meanerrorHardCutFP[trt] <- mean(tmp.err.HardCutFP) / (ntaxa - ncore)
  sim.df$meanerrorPropReadFP[trt] <- mean(tmp.err.PropReadFP) / (ntaxa - ncore)
  sim.df$meanerrorPropRepsFP[trt] <- mean(tmp.err.PropRepsFP) / (ntaxa - ncore)
  sim.df$meanerrorPropReadRepsFP[trt] <- mean(tmp.err.PropReadRepsFP) / (ntaxa - ncore)
}
```

```{r}
myPalette <- c("#800000", "#ffffff", "#C2D1E8", "#678BC6", "#0b45a3")
myPalette2 <- c("#ffffff", "#C2D1E8", "#678BC6", "#0b45a3")
myPalette3 <- c("#ffffff", "#C2D1E8")

df <- data.frame(x = c(1,2,3,4,5), y = c(1,2,3,4,5))

ggplot(df, aes(x = as.factor(x), y = y)) +
  geom_bar(stat = "identity", aes(fill = as.factor(x))) +
  scale_fill_manual(values = myPalette)
```

# True Positive
```{r}
sim.df_gg <- sim.df %>%
  select(1:8) %>%
  gather(method, prob, c(5:8)) %>%
  mutate(method = factor(method, levels = c("meanerrorHardCutTP",
                                            "meanerrorPropReadTP",  
                                            "meanerrorPropRepsTP",
                                          "meanerrorPropReadRepsTP"), labels = c("Hard Cutoff", "Proportion of Reads", "Proportion of Replicates", "Proportion of Reads and Replicates")))

ggplot(sim.df_gg, aes(x = as.factor(corenoncoreratio), y = as.factor(theta))) +
  geom_tile(aes(fill = prob), colour = "white") +
  facet_wrap(.~method, nrow = 2) +
  theme(legend.text=element_text(size=2)) +
  theme_minimal() +
  scale_fill_gradientn(colours = myPalette2, values = scales::rescale(c(-1000, -500, -25,0, 100, 500, 1000))) +
  geom_text(aes(label = round(sim.df_gg$prob, digits = 2))) +
   labs(x = expression(paste("(", pi, " core / ", pi, " non-core)")),
       y = expression(paste(theta)),
       fill = "Mean Error") +
  theme(strip.text.x = element_text(size = 15))

ggsave("figure2.png", width = 9, height  = 8, dpi = 300)
```

# False Positive
```{r}
sim.df_gg <- sim.df %>%
  select(1,2,3,4,9,10,11,12) %>%
  gather(method, prob, c(5:8)) %>%
  mutate(method = factor(method, levels = c("meanerrorHardCutFP",
                                            "meanerrorPropReadFP",  
                                            "meanerrorPropRepsFP",
                                          "meanerrorPropReadRepsFP"), labels = c("Hard Cutoff", "Proportion of Reads", "Proportion of Replicates", "Proportion of Reads and Replicates")))

ggplot(sim.df_gg, aes(x = as.factor(corenoncoreratio), y = as.factor(theta))) +
  geom_tile(aes(fill = prob), colour = "white") +
  facet_wrap(.~method, nrow = 2) +
  theme(legend.text=element_text(size=2)) +
  theme_minimal() +
  scale_fill_gradientn(colours = myPalette2, values = scales::rescale(c(-1000, -500, -25,0, 100, 500, 1000))) +
  geom_text(aes(label = round(sim.df_gg$prob, digits = 2))) +
   labs(x = expression(paste("(", pi, " core / ", pi, " non-core)")),
       y = expression(paste(theta)),
       fill = "Mean Error") +
  theme(strip.text.x = element_text(size = 15))

ggsave("figure3.png", width = 9, height  = 8, dpi = 300)
```


# TP - FP
```{r}
sim.df_gg <- sim.df %>%
  mutate(HardCutoff = (sim.df$meanerrorHardCutTP - sim.df$meanerrorHardCutFP),
         PropReads = (sim.df$meanerrorPropReadTP - sim.df$meanerrorPropReadFP),
         PropReps = (sim.df$meanerrorPropRepsTP - sim.df$meanerrorPropRepsFP),
         PropReadReps = (sim.df$meanerrorPropReadRepsTP - sim.df$meanerrorPropReadRepsFP)) %>%
  select(1,2,3,4,13,14,15,16) %>%
  gather(method, prob, c(5:8)) %>%
  mutate(method = factor(method, levels = c("HardCutoff",
                                            "PropReads",  
                                            "PropReps",
                                          "PropReadReps"), labels = c("Hard Cutoff", "Proportion of Reads", "Proportion of Replicates", "Proportion of Reads and Replicates")))

ggplot(sim.df_gg, aes(x = as.factor(corenoncoreratio), y = as.factor(theta))) +
  geom_tile(aes(fill = prob), colour = "white") +
  facet_wrap(.~method, nrow = 2) +
  theme(legend.text=element_text(size=2)) +
  theme_minimal() +
  scale_fill_gradientn(colours = myPalette, values = scales::rescale(c(-0.5, 0, .5, .75, 1))) +
  geom_text(aes(label = round(sim.df_gg$prob, digits = 2))) +
   labs(x = expression(paste("(", pi, " core / ", pi, " non-core)")),
       y = expression(paste(theta)),
       fill = "Mean Error") +
  theme(strip.text.x = element_text(size = 15))

ggsave("figure4.png", width = 9, height  = 8, dpi = 300)
```

